---
title: "Model 1 — XGBoost"
subtitle: "Gradient-boosted trees, the Kaggle workhorse"
---

> **Where we are:** First of six model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). All six follow the *same*
> structure — recipe → spec → workflow → tune → finalize → evaluate → save — so
> once you've read this one, the others are quick to skim for the differences.
> The features come from the [EDA](01-eda.qmd) and the shared `make_recipe()`.
> Remember the project-wide convention: **we model `log(SalePrice)`**, because
> Kaggle scores this competition as RMSE on the log scale (see the
> [master document](../index.qmd)).

## Why XGBoost first?

Gradient-boosted decision trees are the default first reach on tabular Kaggle
problems. They handle mixed numeric/categorical features, capture interactions
(quality × size, neighborhood × age) automatically, are robust to monotone
transforms, and tune to strong accuracy with modest effort. We tune four of the
most impactful knobs.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

# Project setup: tidyverse + tidymodels, seed = 42, paths, parallel backend.
source(here::here("R", "00_setup.R"))
# Shared feature engineering: defines make_recipe(data).
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(vip)         # variable importance plots
library(xgboost)     # the engine
```

```{r}
#| label: load-data
#| message: false

# Model the LOG of SalePrice — this is the scale Kaggle scores on, so training
# with RMSE here directly optimises the leaderboard metric.
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## The recipe

The shared recipe encodes everything we justified in the [EDA](01-eda.qmd):
impute `LotFrontage`, Yeo-Johnson the skewed areas, lump rare factor levels,
dummy-encode, drop the `Id`, and append principal components.

```{r}
#| label: recipe
xgb_recipe <- make_recipe(train)
xgb_recipe
```

## The model specification

We mark the four tuning parameters with `tune()`. tidymodels fills them in
during the grid search. Note `set_mode("regression")`.

```{r}
#| label: spec
xgb_spec <- boost_tree(
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n      = tune()
) |>
  set_engine("xgboost") |>
  set_mode("regression")

xgb_spec
```

## The workflow

A `workflow` bundles the recipe and the model so resampling applies the *exact*
same preprocessing to every fold — no leakage.

```{r}
#| label: workflow
xgb_wf <- workflow() |>
  add_recipe(xgb_recipe) |>
  add_model(xgb_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
# Stratify on the (log) outcome so each fold spans the full price range.
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
# Predictors used to finalize data-dependent params (num_comp upper bound).
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

xgb_params <- extract_parameter_set_dials(xgb_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 15L)))

set.seed(42)
xgb_grid <- grid_space_filling(xgb_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false

# Score every candidate on RMSE (the competition metric on the log scale) and
# R-squared (interpretable goodness-of-fit).
xgb_metrics <- metric_set(rmse, rsq)

set.seed(42)
xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = xgb_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)
```

```{r}
#| label: save-res
saveRDS(xgb_tuned, here("models", "xgboost_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid for each hyperparameter (autoplot)."
#| fig-width: 9
#| fig-height: 6

autoplot(xgb_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
# The top candidates by RMSE (lower is better).
show_best(xgb_tuned, metric = "rmse", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_xgb <- select_best(xgb_tuned, metric = "rmse")
best_xgb

final_xgb_wf <- finalize_workflow(xgb_wf, best_xgb)

# Fit the finalized workflow on ALL the training data — the model we ship.
set.seed(42)
xgb_fit <- fit(final_xgb_wf, data = train)
```

## Cross-validated performance

The fairest estimate of out-of-sample skill is the CV score of the chosen
configuration — not the fit on the full training data (which overfits by
construction).

```{r}
#| label: cv-metrics
collect_metrics(xgb_tuned) |>
  semi_join(best_xgb, by = names(best_xgb)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
The `rmse` here is on the **log(SalePrice)** scale — the same units Kaggle
reports on the leaderboard. These CV metrics feed the results table on the
[master document](../index.qmd) automatically.
:::

### Observed vs predicted (out-of-fold)

Built from the held-out CV predictions for the best configuration — so it
reflects generalization, not memorization. Points on the diagonal are perfect.

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5

collect_predictions(xgb_tuned) |>
  filter(.config == best_xgb$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[1]) +
  labs(title = "XGBoost — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Top features by XGBoost gain. Quality and size dominate."
#| fig-width: 7
#| fig-height: 5

xgb_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = house_pal[1])) +
  labs(title = "XGBoost variable importance") +
  theme_minimal()
```

`OverallQual`, `GrLivArea`, the engineered `TotalSF`, and neighborhood/age
features sit at the top — exactly the signal the [EDA](01-eda.qmd) surfaced.

## Save the fitted workflow

```{r}
#| label: save
saveRDS(xgb_fit, file = here("models", "xgboost_fit.rds"))
```

## Submission

Generate this model's leaderboard CSV with the **parameterized**
[submission document](submission.qmd):

```bash
quarto render docs/submission.qmd -P model:xgboost
```

That loads `models/xgboost_fit.rds`, predicts on the test set, **exponentiates**
back to dollars, and writes `models/submission_xgboost.csv` (columns `Id`,
`SalePrice`). See the [Kaggle setup guide](setup-kaggle.md) for authentication.

## Next steps

- Compare against the **[Random Forest](03-random-forest.qmd)** — a robust,
  almost-no-tuning baseline.
- Then the **[Neural Net](04-neural-net.qmd)**, **[Elastic Net](06-elasticnet.qmd)**,
  **[SVM](07-svm.qmd)** and **[MARS](08-mars.qmd)**.
- Record this model's CV scores in the **[master results table](../index.qmd)**.
