---
title: "Model 2 — Random Forest"
subtitle: "Bagged trees via ranger — a robust, low-tuning baseline"
---

> **Where we are:** Second of six model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). Same structure as
> [XGBoost](02-xgboost.qmd): recipe → spec → workflow → tune → finalize →
> evaluate → save. We model `log(SalePrice)` and score on RMSE.

## Why a random forest?

Where boosting builds trees sequentially to correct residuals, a random forest
*bags* hundreds of decorrelated deep trees and averages them. It is famously
hard to overfit, needs almost no tuning (really just `mtry`), and gives a strong
baseline plus a permutation-importance read on which features matter.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(vip)
library(ranger)   # the engine
```

```{r}
#| label: load-data
#| message: false
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## Recipe, spec, workflow

```{r}
#| label: recipe-spec
rf_recipe <- make_recipe(train)

# 1000 trees fixed (more is just slower, not better once stable); tune mtry and
# min_n. importance = "permutation" so we can plot variable importance.
rf_spec <- rand_forest(
  mtry  = tune(),
  min_n = tune(),
  trees = 1000
) |>
  set_engine("ranger", importance = "permutation") |>
  set_mode("regression")

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
# mtry's upper bound depends on the number of predictors, so finalize against
# the baked training data.
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

rf_params <- extract_parameter_set_dials(rf_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 15L)))

set.seed(42)
rf_grid <- grid_space_filling(rf_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false
rf_metrics <- metric_set(rmse, rsq)

set.seed(42)
rf_tuned <- tune_grid(
  rf_wf,
  resamples = folds,
  grid      = rf_grid,
  metrics   = rf_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)

saveRDS(rf_tuned, here("models", "rf_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 6
autoplot(rf_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
show_best(rf_tuned, metric = "rmse", n = 5)
```

## Finalize and fit

```{r}
#| label: finalize
best_rf <- select_best(rf_tuned, metric = "rmse")
best_rf

final_rf_wf <- finalize_workflow(rf_wf, best_rf)

set.seed(42)
rf_fit <- fit(final_rf_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(rf_tuned) |>
  semi_join(best_rf, by = names(best_rf)) |>
  select(.metric, mean, std_err, n)
```

### Observed vs predicted (out-of-fold)

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5
collect_predictions(rf_tuned) |>
  filter(.config == best_rf$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[1]) +
  labs(title = "Random Forest — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Top features by permutation importance."
#| fig-width: 7
#| fig-height: 5
rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = house_pal[2])) +
  labs(title = "Random forest variable importance") +
  theme_minimal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(rf_fit, file = here("models", "rf_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:rf
```

## Next steps

- Continue to the **[Neural Net](04-neural-net.qmd)**, which leans on the
  normalized/PCA recipe.
- Back to the **[master results table](../index.qmd)**.
