---
title: "Model 3 — Neural Net"
subtitle: "A small single-hidden-layer MLP via nnet"
---

> **Where we are:** Third of six model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). Same structure as the others;
> we model `log(SalePrice)` and score on RMSE. The neural net is the model that
> most needs the recipe's `step_normalize()` + `step_pca()` — MLPs are sensitive
> to input scale in a way tree ensembles are not.

## Why a neural net here?

A single-hidden-layer MLP learns a smooth nonlinear function of the (normalized,
decorrelated) principal components. On a well-featured tabular problem it rarely
beats gradient boosting, but it makes *different* errors — which is exactly what
makes it a valuable member of the [stacked ensemble](05-stack.qmd) later.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(nnet)   # the engine
```

```{r}
#| label: load-data
#| message: false
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## Recipe, spec, workflow

The shared recipe already normalizes all numeric predictors and appends PCs, so
the MLP sees well-behaved inputs. We tune the hidden-layer width and the weight
decay `penalty`, and lean on `num_comp` to control input dimensionality.

```{r}
#| label: recipe-spec
nnet_recipe <- make_recipe(train)

nnet_spec <- mlp(
  hidden_units = tune(),
  penalty      = tune(),
  epochs       = 300
) |>
  # MaxNWts must be large enough for (many predictors) x (hidden units); trace
  # off to keep the tuning log quiet. linout is set automatically in regression.
  set_engine("nnet", MaxNWts = 10000, trace = FALSE) |>
  set_mode("regression")

nnet_wf <- workflow() |>
  add_recipe(nnet_recipe) |>
  add_model(nnet_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

nnet_params <- extract_parameter_set_dials(nnet_wf) |>
  finalize(preds) |>
  update(
    hidden_units = hidden_units(range = c(2L, 15L)),
    num_comp     = num_comp(range = c(2L, 15L))
  )

set.seed(42)
nnet_grid <- grid_space_filling(nnet_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false
nnet_metrics <- metric_set(rmse, rsq)

set.seed(42)
nnet_tuned <- tune_grid(
  nnet_wf,
  resamples = folds,
  grid      = nnet_grid,
  metrics   = nnet_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)

saveRDS(nnet_tuned, here("models", "nnet_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 6
autoplot(nnet_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
show_best(nnet_tuned, metric = "rmse", n = 5)
```

## Finalize and fit

```{r}
#| label: finalize
best_nnet <- select_best(nnet_tuned, metric = "rmse")
best_nnet

final_nnet_wf <- finalize_workflow(nnet_wf, best_nnet)

set.seed(42)
nnet_fit <- fit(final_nnet_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(nnet_tuned) |>
  semi_join(best_nnet, by = names(best_nnet)) |>
  select(.metric, mean, std_err, n)
```

### Observed vs predicted (out-of-fold)

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5
collect_predictions(nnet_tuned) |>
  filter(.config == best_nnet$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[3]) +
  labs(title = "Neural Net — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(nnet_fit, file = here("models", "nnet_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:nnet
```

## Next steps

- Continue to the interpretable **[Elastic Net](06-elasticnet.qmd)** linear
  baseline, then **[SVM](07-svm.qmd)** and **[MARS](08-mars.qmd)**.
- Back to the **[master results table](../index.qmd)**.
