---
title: "Model 7 — Stacked Ensemble"
subtitle: "Blending six base models (XGBoost, RF, Neural Net, Elastic Net, SVM, MARS) with stacks"
---

> **Where we are:** The capstone of the
> [House Prices — Tidymodels Tour](../index.qmd). The six base-model documents
> ([XGBoost](02-xgboost.qmd), [Random Forest](03-random-forest.qmd),
> [Neural Net](04-neural-net.qmd), [Elastic Net](06-elasticnet.qmd),
> [SVM](07-svm.qmd), [MARS](08-mars.qmd)) each saved their *tuning results*
> (every candidate model and its out-of-fold predictions) to
> `models/<model>_res.rds`. Here we feed all of those candidates to
> [`stacks`](https://stacks.tidymodels.org/), which learns a regularised linear
> blend over them — usually beating any single model, because the members make
> different mistakes.

## Why stack?

Each model family has a different inductive bias: gradient boosting chases
interactions, the random forest averages decorrelated trees, the neural net and
SVM learn smooth functions of the principal components, the elastic net is
linear, MARS is piecewise-linear. A stack fits a *meta-learner* (penalised
linear regression) on the members' out-of-fold predictions, keeping only the
candidates that pull their weight. Everything stays on the `log(SalePrice)`
scale until the final submission.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(stacks)
# The member engines must be attached so fit_members() can refit them.
library(ranger)
library(xgboost)
library(nnet)
library(glmnet)
library(kernlab)
library(earth)
```

## Load the tuning results

```{r}
#| label: load-candidates
# Each *_res.rds is a tune_grid result saved with control_grid(save_pred = TRUE,
# save_workflow = TRUE) — exactly what stacks needs to assemble candidates.
xgb_res  <- readRDS(here("models", "xgboost_res.rds"))
rf_res   <- readRDS(here("models", "rf_res.rds"))
nnet_res <- readRDS(here("models", "nnet_res.rds"))
en_res   <- readRDS(here("models", "elasticnet_res.rds"))
svm_res  <- readRDS(here("models", "svm_res.rds"))
mars_res <- readRDS(here("models", "mars_res.rds"))
```

## Build, blend, and fit the ensemble

```{r}
#| label: build-stack
#| message: false
#| warning: false

set.seed(42)
house_stack <- stacks() |>
  add_candidates(xgb_res) |>
  add_candidates(rf_res) |>
  add_candidates(nnet_res) |>
  add_candidates(en_res) |>
  add_candidates(svm_res) |>
  add_candidates(mars_res) |>
  blend_predictions() |>     # LASSO over candidates -> keeps the useful ones
  fit_members()              # refit the retained members on all training data

house_stack
```

### Which members survived the blend?

```{r}
#| label: weights
#| fig-cap: "Blended stacking coefficients — the members the meta-learner kept."
#| fig-width: 8
#| fig-height: 5
autoplot(house_stack, type = "weights") +
  scale_colour_manual(values = grDevices::colorRampPalette(house_pal)(10)) +
  scale_fill_manual(values = grDevices::colorRampPalette(house_pal)(10))
```

The boosted trees and elastic net usually earn the largest weights, with the SVM
and neural net contributing diverse corrections — different biases, combined.

## Save the fitted ensemble

```{r}
#| label: save
saveRDS(house_stack, here("models", "stack_fit.rds"))
```

## Build the Kaggle submission

```{r}
#| label: submit
test <- read_csv(here("data", "processed", "test.csv"), show_col_types = FALSE)

# Predictions are on the log scale (we trained on log(SalePrice)); exponentiate
# back to dollars for the leaderboard.
stack_pred <- predict(house_stack, new_data = test)

submission <- tibble(
  Id        = test$Id,
  SalePrice = exp(stack_pred$.pred)
)
write_csv(submission, here("models", "submission_stack.csv"))

summary(submission$SalePrice)
```

Upload it the same way as any other model:

```bash
kaggle competitions submit -c home-data-for-ml-course -f models/submission_stack.csv \
  -m "stacked ensemble (xgb + rf + nnet + glmnet + svm + mars) via stacks"
```

## Next steps

- The parameterized [submission doc](submission.qmd) also accepts
  `-P model:stack`.
- Compare the stack's leaderboard score against the individual models on the
  [master document](../index.qmd) — does the blend actually win on the held-out
  Kaggle test set, or only in cross-validation?
- See the [Kaggle setup guide](setup-kaggle.md) for authentication.
