---
title: "Model 4 — Elastic Net"
subtitle: "Penalised linear regression via glmnet — interpretable baseline"
---

> **Where we are:** Fourth of six model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). Same structure as the others;
> we model `log(SalePrice)` and score on RMSE.

## Why elastic net?

A regularised linear model is the classic, interpretable regression baseline —
and on the log scale, with the Yeo-Johnson-transformed features from the recipe,
it is genuinely competitive on Ames. `glmnet` blends **ridge** (L2) and
**lasso** (L1) penalties: `penalty` sets the total shrinkage and `mixture` the
L1/L2 balance (1 = pure lasso, 0 = pure ridge). Lasso also does automatic
feature selection, which is handy given ~200 dummy columns.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(vip)
library(glmnet)   # the engine
```

```{r}
#| label: load-data
#| message: false
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## Recipe, spec, workflow

```{r}
#| label: recipe-spec
en_recipe <- make_recipe(train)

en_spec <- linear_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet") |>
  set_mode("regression")

en_wf <- workflow() |>
  add_recipe(en_recipe) |>
  add_model(en_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

en_params <- extract_parameter_set_dials(en_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 15L)))

set.seed(42)
en_grid <- grid_space_filling(en_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false
en_metrics <- metric_set(rmse, rsq)

set.seed(42)
en_tuned <- tune_grid(
  en_wf,
  resamples = folds,
  grid      = en_grid,
  metrics   = en_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)

saveRDS(en_tuned, here("models", "elasticnet_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 6
autoplot(en_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
show_best(en_tuned, metric = "rmse", n = 5)
```

## Finalize and fit

```{r}
#| label: finalize
best_en <- select_best(en_tuned, metric = "rmse")
best_en

final_en_wf <- finalize_workflow(en_wf, best_en)

set.seed(42)
en_fit <- fit(final_en_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(en_tuned) |>
  semi_join(best_en, by = names(best_en)) |>
  select(.metric, mean, std_err, n)
```

### Observed vs predicted (out-of-fold)

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5
collect_predictions(en_tuned) |>
  filter(.config == best_en$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[4]) +
  labs(title = "Elastic Net — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Largest standardised coefficients at the tuned penalty."
#| fig-width: 7
#| fig-height: 5
en_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = house_pal[4])) +
  labs(title = "Elastic net — top coefficients") +
  theme_minimal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(en_fit, file = here("models", "elasticnet_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:elasticnet
```

## Next steps

- Continue to the **[SVM](07-svm.qmd)** and **[MARS](08-mars.qmd)** models.
- Back to the **[master results table](../index.qmd)**.
