---
title: "Model 5 — Support Vector Machine (RBF)"
subtitle: "A nonlinear kernel regressor via kernlab"
---

> **Where we are:** Fifth of six model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). Same structure as the others;
> we model `log(SalePrice)` and score on RMSE.

## Why an SVM?

Support vector regression with a radial-basis kernel draws a smooth, nonlinear
response surface over the normalized principal components. Its errors are
structurally different from the trees' and the linear model's — diversity that
pays off in the [stack](05-stack.qmd). It needs scaled inputs, which the shared
recipe already provides. We tune the regularisation `cost` and the kernel width
`rbf_sigma`.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(kernlab)   # the engine
```

```{r}
#| label: load-data
#| message: false
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## Recipe, spec, workflow

```{r}
#| label: recipe-spec
svm_recipe <- make_recipe(train)

svm_spec <- svm_rbf(
  cost      = tune(),
  rbf_sigma = tune()
) |>
  set_engine("kernlab") |>
  set_mode("regression")

svm_wf <- workflow() |>
  add_recipe(svm_recipe) |>
  add_model(svm_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

svm_params <- extract_parameter_set_dials(svm_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 15L)))

set.seed(42)
svm_grid <- grid_space_filling(svm_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false
svm_metrics <- metric_set(rmse, rsq)

set.seed(42)
svm_tuned <- tune_grid(
  svm_wf,
  resamples = folds,
  grid      = svm_grid,
  metrics   = svm_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)

saveRDS(svm_tuned, here("models", "svm_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 6
autoplot(svm_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
show_best(svm_tuned, metric = "rmse", n = 5)
```

## Finalize and fit

```{r}
#| label: finalize
best_svm <- select_best(svm_tuned, metric = "rmse")
best_svm

final_svm_wf <- finalize_workflow(svm_wf, best_svm)

set.seed(42)
svm_fit <- fit(final_svm_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(svm_tuned) |>
  semi_join(best_svm, by = names(best_svm)) |>
  select(.metric, mean, std_err, n)
```

### Observed vs predicted (out-of-fold)

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5
collect_predictions(svm_tuned) |>
  filter(.config == best_svm$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[5]) +
  labs(title = "SVM (RBF) — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(svm_fit, file = here("models", "svm_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:svm
```

## Next steps

- Finish the base models with **[MARS](08-mars.qmd)**, then blend everything in
  the **[Stacked Ensemble](05-stack.qmd)**.
- Back to the **[master results table](../index.qmd)**.
