---
title: "Model 6 — MARS"
subtitle: "Multivariate Adaptive Regression Splines via earth"
---

> **Where we are:** Last of six base model documents in the
> [House Prices — Tidymodels Tour](../index.qmd). Same structure as the others;
> we model `log(SalePrice)` and score on RMSE.

## Why MARS?

MARS fits the response as a sum of **hinge functions** — piecewise-linear splines
with automatically chosen knots — and can multiply them for interactions. That
makes it naturally suited to housing data, where a feature's effect often
*changes slope* at a threshold (e.g. living area matters more up to a point). It
is nonlinear yet interpretable, and diverse from the trees. We tune the number
of retained terms and the interaction degree.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(vip)
library(earth)   # the engine
```

```{r}
#| label: load-data
#| message: false
train <- readr::read_csv(here("data", "processed", "train.csv"),
                         show_col_types = FALSE) |>
  mutate(SalePrice = log(SalePrice))
```

## Recipe, spec, workflow

```{r}
#| label: recipe-spec
mars_recipe <- make_recipe(train)

mars_spec <- mars(
  num_terms   = tune(),
  prod_degree = tune()
) |>
  set_engine("earth") |>
  set_mode("regression")

mars_wf <- workflow() |>
  add_recipe(mars_recipe) |>
  add_model(mars_spec)
```

## Tuning with 10-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = SalePrice)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-SalePrice, -Id)

mars_params <- extract_parameter_set_dials(mars_wf) |>
  finalize(preds) |>
  update(
    num_terms = num_terms(range = c(5L, 50L)),
    num_comp  = num_comp(range = c(2L, 15L))
  )

set.seed(42)
mars_grid <- grid_space_filling(mars_params, size = 20)
```

```{r}
#| label: tune
#| message: false
#| warning: false
mars_metrics <- metric_set(rmse, rsq)

set.seed(42)
mars_tuned <- tune_grid(
  mars_wf,
  resamples = folds,
  grid      = mars_grid,
  metrics   = mars_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE)
)

saveRDS(mars_tuned, here("models", "mars_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "RMSE across the tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 6
autoplot(mars_tuned, metric = "rmse")
```

```{r}
#| label: tune-table
show_best(mars_tuned, metric = "rmse", n = 5)
```

## Finalize and fit

```{r}
#| label: finalize
best_mars <- select_best(mars_tuned, metric = "rmse")
best_mars

final_mars_wf <- finalize_workflow(mars_wf, best_mars)

set.seed(42)
mars_fit <- fit(final_mars_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(mars_tuned) |>
  semi_join(best_mars, by = names(best_mars)) |>
  select(.metric, mean, std_err, n)
```

### Observed vs predicted (out-of-fold)

```{r}
#| label: obs-pred
#| fig-cap: "Out-of-fold predicted vs observed log(SalePrice)."
#| fig-width: 5.5
#| fig-height: 5
collect_predictions(mars_tuned) |>
  filter(.config == best_mars$.config) |>
  ggplot(aes(x = SalePrice, y = .pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(alpha = 0.35, colour = house_pal[6]) +
  labs(title = "MARS — CV predicted vs observed",
       x = "Observed log(SalePrice)", y = "Predicted log(SalePrice)") +
  coord_equal()
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Features retained in the MARS model, by importance."
#| fig-width: 7
#| fig-height: 5
mars_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = house_pal[6])) +
  labs(title = "MARS variable importance") +
  theme_minimal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(mars_fit, file = here("models", "mars_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:mars
```

## Next steps

- Blend all six base models in the **[Stacked Ensemble](05-stack.qmd)**.
- Back to the **[master results table](../index.qmd)**.
