---
title: "Building a Kaggle Submission"
subtitle: "One parameterized document, seven models, leaderboard-ready CSVs"
params:
  model: "xgboost"
execute:
  freeze: false   # parameterized: must re-run per -P model:, never serve a cached model
---

> **Where we are:** The final stop on the
> [House Prices — Tidymodels Tour](../index.qmd). Each model document fits a
> workflow and saves it to `models/<model>_fit.rds`. This document is
> **parameterized**: tell it which model, and it loads that fit, predicts on the
> test set, **exponentiates** predictions back to dollars, and writes a
> Kaggle-ready CSV.

## How to render this document per model

This `.qmd` has a `model` parameter (default `"xgboost"`). Override it on the
command line with `-P model:<value>`:

```bash
quarto render docs/submission.qmd -P model:xgboost
quarto render docs/submission.qmd -P model:rf
quarto render docs/submission.qmd -P model:nnet
quarto render docs/submission.qmd -P model:elasticnet
quarto render docs/submission.qmd -P model:svm
quarto render docs/submission.qmd -P model:mars
quarto render docs/submission.qmd -P model:stack
```

Each render reads the matching `models/<model>_fit.rds`, writes
`models/submission_<model>.csv`, and prints the exact upload command. You must
have fitted (rendered) the corresponding model document first so the `.rds`
exists.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
library(tidyverse)
library(tidymodels)
library(here)
```

## Resolve and validate the requested model

```{r}
#| label: resolve-model
requested_model <- params$model

# This single lookup table is the source of truth for which models are submittable.
model_paths <- c(
  xgboost    = here::here("models", "xgboost_fit.rds"),
  rf         = here::here("models", "rf_fit.rds"),
  nnet       = here::here("models", "nnet_fit.rds"),
  elasticnet = here::here("models", "elasticnet_fit.rds"),
  svm        = here::here("models", "svm_fit.rds"),
  mars       = here::here("models", "mars_fit.rds"),
  stack      = here::here("models", "stack_fit.rds")
)

if (!requested_model %in% names(model_paths)) {
  stop(
    "Unknown model '", requested_model, "'. ",
    "Choose one of: ", paste(names(model_paths), collapse = ", "), ". ",
    "Example: quarto render docs/submission.qmd -P model:rf",
    call. = FALSE
  )
}

fit_path <- model_paths[[requested_model]]

if (!file.exists(fit_path)) {
  stop(
    "Fitted model not found at '", fit_path, "'. ",
    "Render the matching model document first (e.g. quarto render ",
    "docs/02-xgboost.qmd) so the .rds is saved.",
    call. = FALSE
  )
}

message("Building submission for model: ", requested_model)
message("Loading fitted workflow from: ", fit_path)
```

We are building a submission for **`r requested_model`**.

## Load the model and the test data

```{r}
#| label: load
model_fit <- readRDS(fit_path)

# Use the PROCESSED test set (engineered features added in R/03_engineer.R), so
# the columns match what every model/recipe expects. Id keys the submission.
test <- read_csv(here("data", "processed", "test.csv"), show_col_types = FALSE)

glimpse(test)
```

## Predict and assemble the submission

```{r}
#| label: predict
# Every model was trained on log(SalePrice), so predictions are on the log
# scale. predict() runs the workflow's recipe on the test data, then the model.
preds <- predict(model_fit, new_data = test)

# Kaggle requires EXACTLY two columns: Id and SalePrice (in dollars). We
# exponentiate the log-scale prediction back to dollars.
submission <- tibble(
  Id        = test$Id,
  SalePrice = exp(preds$.pred)
)

# Sanity checks: right number of rows, positive prices, no missing predictions.
stopifnot(
  nrow(submission) == nrow(test),
  all(submission$SalePrice > 0),
  !anyNA(submission$SalePrice)
)

head(submission, 10)
```

## Write the CSV

```{r}
#| label: write
out_path <- here("models", paste0("submission_", requested_model, ".csv"))
write_csv(submission, out_path)
message("Wrote ", nrow(submission), " predictions to: ", out_path)
```

A quick summary of the predicted prices (sanity check against the ~$34k–$755k
range seen in the [EDA](01-eda.qmd)):

```{r}
#| label: summary
summary(submission$SalePrice)
```

## Submit to Kaggle

```{r}
#| label: kaggle-cmd
#| echo: false
#| results: asis

cmd <- sprintf(
  'kaggle competitions submit -c home-data-for-ml-course -f models/submission_%s.csv -m "%s via tidymodels"',
  requested_model, requested_model
)
cat("```bash\n", cmd, "\n```\n", sep = "")
```

::: {.callout-note}
## First time? Authenticate first
The `kaggle` CLI needs your API token (`~/.kaggle/kaggle.json`) before any
`submit` or `download` will work. The
[Kaggle setup guide](setup-kaggle.md) walks through creating and installing it.
:::

::: {.callout-tip}
## The leaderboard metric
This competition is scored on **RMSE between the log of the predicted and the
log of the observed SalePrice**. Because we train on `log(SalePrice)`, each
model's CV `rmse` (on the [master results table](../index.qmd)) is already in
the same units as the leaderboard — a rare, satisfying alignment.
:::

## Next steps

- Compare each model's leaderboard score against its **CV score** recorded in
  the [master results table](../index.qmd) — divergence is a lesson in
  generalization.
- Re-render for the other models with `-P model:<name>`.
- Need auth help? See the **[Kaggle setup guide](setup-kaggle.md)**.
